---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data

```{r}
library(dplyr, quietly = TRUE, warn.conflicts = FALSE)
if(!file.exists('activity.csv')) {
  unzip('activity.zip')
}
activity <- read.csv('activity.csv')

```

## What is mean total number of steps taken per day?

```{r}
steps_per_date <- aggregate(steps ~ date, data=activity, sum, na.rm=TRUE)
```

```{r}
hist(steps_per_date$steps, 
     main = "Histogram of steps/day",
     xlab = "Number of steps")
```

```{r}
mean_steps_daily <- mean(steps_per_date$steps)
median_steps_daily <- median(steps_per_date$steps)
print(paste("Mean steps per day: ", mean_steps_daily))
print(paste("Median steps per day: ", median_steps_daily))
```

## What is the average daily activity pattern?

```{r}
activity_pattern_agg <- aggregate(steps ~ interval, 
                                  data = activity,
                                  mean,
                                  na.rm = TRUE)
plot(activity_pattern_agg$interval, 
     activity_pattern_agg$steps, 
     type='l',
     main='Avg. steps per time interval',
     xlab='Interval',
     ylab='Avg. steps')
```

```{r}
max_rows <- activity_pattern_agg[which.max(activity_pattern_agg$steps),]
print(paste("Interval with maximum average steps: ", max_rows$interval))
```

## Imputing missing values

The `date` and `interval` columns are complete, however the `steps` column is missing some values.

```{r}
# Confirming that we have no missing date or interval columns.
missing_dates <- sum(is.na(activity$date))
missing_intervals <- sum(is.na(activity$interval))

# We are missing some steps values though.
missing_steps <- sum(is.na(activity$steps))
print(paste("Number of missing step values: ", missing_steps))
```

To fill in the missing values, we will populate the `NA` values with the medians for that time interval.

```{r}
imputed_activity <- activity %>%
  mutate(steps=as.double(steps)) %>%
  inner_join(activity_pattern_agg, by="interval") %>%
  mutate(steps=coalesce(steps.x, steps.y)) %>%
  select(date, interval, steps)

steps_per_date_imputed <- aggregate(steps ~ date, 
                                    data = imputed_activity, 
                                    sum)
hist(steps_per_date_imputed$steps,
     main = "Histogram of steps/day", 
     xlab = "Number of steps")

```

Clearly, imputing the missing values alters the frequency of the daily step counts.  Of note is the fact that the distribution remains roughly the same.

```{r}
mean_steps_daily_imputed <- mean(steps_per_date_imputed$steps)
median_steps_daily_imputed <- median(steps_per_date_imputed$steps)
print(paste("Mean steps per day (imputed): ", mean_steps_daily_imputed))
print(paste("Median steps per day (imputed): ", median_steps_daily_imputed))
```

Imputing the missing value with this method does not cause any significant changes to the daily mean/median values.

## Are there differences in activity patterns between weekdays and weekends?
